Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-23 manuscript.Rmd
background 1

Recent debates about whether adults undergo intrapersonal cultural change emerged in part because sociological models of cultural change at the aggregate level tend to implicitly invoke one of two models of individual behavior. The first, what Kiley and Vaisey [-@kiley2020] call a “Settled Dispositions Model” (SDM), assumes that peoples’ personal culture is relatively fixed by the time they are adults. While they might make temporary changes in their declarative culture as their social environments provide new considerations, this model assumes that they return to a settled baseline over a (relatively) short time frame. This model of behavior is frequently invoked in models of cultural change that suggest people are strongly imprinted by the “past conditions of production” of socialization, such as Bourdieusian practice theories [@bourdieu1990], cohort replacement theories of aggregate change [@mannheim1952; @ryder1965], and control theories in social psychology [@smith-lovin1988; @robinson2007].

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people make new cultural meanings, and therefore durably change their personal culture, as they encounter new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies models of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018], among others. It is also implicit in most studies of whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

There is no reason to think that only one of these models is “correct” in all times and places. A population observed over some time frame contains a mix of people who are changing and people who are not (or some people making very small or imperceptible changes in culture and others making perceptible changes). Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood, typically viewed as a “formative period” for personal cultural development, is likely characterized by higher rates of active updating, while middle age and later life are potentially characterized more by settled dispositions [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while low-salience, established issues are characterized by stability.

Comparing which of these models better fit a broad range of questions from the General Social Survey’s rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the AUM was preferred for the majority of items, the total amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they showed no evidence of durable change. Questions with more evidence of durable change included highly salient issues like gay marriage and “public” issues like partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than the rest of the population. Overall, the researchers concluded that “results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other findings that cohort replacement plays a larger role than period effects in explaining differences in personal culture [@vaisey2016].

However, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying durable individual change across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021], as a result of social experiences. Similarly, there are longitudinal studies showing that cues from political elites can change a person’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in a person’s close contacts and acquaintances can change their attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023] and proposed a “Life Course Adaption Model” (LCAM), which draws on the life course perspective and models personal culture over the course of a panel as a linear trajectory that varies for each respondent. Lersch’s work addresses two shortcomings of the AUM and SDM comparison. First, the AUM as outlined by Kiley and Vaisey assumes that change can only occur as a Markovian process, in which responses to personal culture items at time $t$ are a function only of responses at time $t-1$. But early life experiences might indirectly shape peoples’ personal culture as they move into new social roles or environments, even if they do not affect personal culture right away. For example, experiences in childhood might not affect views about family structure until much later in life when people create their own families. The LCAM allows responses at earlier points than $t-1$ to influence responses at time $t$.

>>>>

Recent debates about whether adults undergo intrapersonal cultural change emerged in part because theories of cultural change at the aggregate level tend to implicitly invoke one of two models of individual behavior. The first, what Kiley and Vaisey [-@kiley2020] call a “Settled Dispositions Model” (SDM), assumes that peoples’ personal culture is relatively fixed by the time they are adults. While they might make temporary changes in their declarative culture in reaction to their environments, this model assumes that people return to a settled baseline over a (relatively) short period of time. This model is frequently invoked in theories of cultural change that suggest people are strongly imprinted by early socialization experiences such as the “past conditions of production” a Bourdieusian practice theory [@bourdieu1990], cohort replacement theories of aggregate change [@mannheim1952; @ryder1965], and control theories in social psychology [@smith-lovin1988; @robinson2007].

The second model summarized by Kiley and Vaisey [-@kiley2020], an “Active Updating Model” (AUM), posits that people continually update their personal culture as they move through life. This model suggests people change their personal culture as they adapt and make new meanings when encountering new social environments, discourses, and information [@gross2009; @swidler2001]. This model underlies, among others, theories of cultural diffusion [@christakis2010], attitude alignment [@dellaposta2020], and polarization [@bail2018] It is also implicit in most studies that ask whether specific experiences, changes in social roles, or political events, affect personal culture [@gelman2021; @visser2004; @slothuus2021].

There is no reason to believe that only one of these two models is “correct” at all times and in all places. A population observed over some period of time contains a mix of people who are noticeably changing and people who are not. Instead, different perspectives argue that each of these ideal-typical models is more operative at different times, for different people, and for different elements of personal culture. For example, adolescence and early adulthood is typically viewed as a “formative period” for personal cultural development and thus characterized by higher rates of active updating, while middle age and later life are potentially characterized more by settled dispositions [@alwin1991; @krosnick1989; @eaton2009]. Similarly, highly salient issues, such as views around gay rights in the 2010s; issues that see substantial elite realignments, such as views around the Vietnam conflict in the 1970s [@zaller1992]; and novel issues of public opinion, such as views around vaccines during the Covid-19 pandemic, might be characterized by active updating, while established issues of low salience are characterized by stability.

Empirically comparing which of these models better fit a broad range of questions from the General Social Survey’s rotating panels, Kiley and Vaisey [-@kiley2020] found limited evidence of durable change. While the AUM was preferred for the majority of items, the total amount of durable change detected on these items was small. A substantial minority of questions (39 percent) favored the SDM, meaning they showed no evidence of durable change. Questions with more evidence of durable change included highly salient issues like gay marriage and “public” issues like partisan identification and religious service attendance. There was also more evidence of durable change among early adults (people ages 18-30) than the rest of the population. Overall, the researchers concluded that “results ultimately suggest that real, persistent attitude change is an uncommon phenomenon among adults'' [@kiley2020: p. 500; see also @vaisey2021]. This lack of durable change is consistent with other findings that cohort replacement plays a larger role than period effects in explaining differences in personal culture [@vaisey2016].

On the other hand, the claim that change is a relatively infrequent phenomenon has been difficult to square with research identifying durable change as a result of social experiences across a number of cultural dimensions, such as morality [@brocic2021], trust [@mewes2021], and concerns about immigration [@kratz2021]. Similarly, there are longitudinal studies showing that cues from political elites can change an individual’s position on specific policy issues [@slothuus2021; @zaller1992] and that changes in their close contacts and acquaintances can change individuals' attitudes on group-related politics [@gelman2021; @dellaposta2018]. Given how often we observe people change their personal culture, it is hard to accept that adults do not change.

Drawing on these findings, Lersch challenged the SDM and AUM as a “needless dichotomy” [-@lersch2023] and proposed a “Life Course Adaption Model” (LCAM) instead, which draws on the life course perspective and models personal culture over the course of a panel as a linear trajectory that varies for each respondent. Lersch’s work addresses two shortcomings of the AUM and SDM comparison. First, the AUM as outlined by Kiley and Vaisey assumes that change can only occur as a Markovian process, in which responses to personal culture items at time $t$ are a function only of responses at time $t-1$. But early life experiences might indirectly shape peoples’ personal culture as they move into new social roles or environments, even if they do not affect personal culture right away. For example, experiences in childhood might not affect views about family structure until much later in life when people create their own families. The LCAM allows responses at earlier points than $t-1$ to influence responses at time $t$.
